---
title: "R Workshop <br> Based on Duke MEM <br> by Mine Cetinkaya-Rundel"
author: "Jan Idziak"
date: "11 September, 2016"
output: ioslides_presentation
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

- Initial data preparation

- Data investigation

- Basic data visualizations in ggplot
    - scatter plot
    - box plot
    - histogram
    - bar chart

- data munging using dplyr

# Data visualization

## Load data 

```{r load-data, warning=FALSE, message=FALSE}
library(caret)
library(dplyr)
library(ggplot2)
library(plotly)
data("GermanCredit")
```

## Data Preparation {.smaller}

```{r eval = FALSE} 
str(GermanCredit)
```

```{r , cache=TRUE}
names(GermanCredit)[1:20]
```

## Data Preparation cd...

```{r , cache=TRUE}
GermanCredit$Sex <- 
  GermanCredit$Personal.Male.Divorced.Seperated +
  GermanCredit$Personal.Male.Single + 
  GermanCredit$Personal.Male.Married.Widowed

GermanCredit$Sex <- ifelse(GermanCredit$Sex, "Male", "Female")
```

## Data Preparation cd...

```{r , cache=TRUE}
GermanCredit$Property <- 
  sapply(1:dim(GermanCredit)[1], function(i)
    if(GermanCredit$Property.RealEstate[i]){
      "RealEstate"
    } else if(GermanCredit$Property.Insurance[i]) {
      "Insurance"
    } else if(GermanCredit$Property.CarOther[i]) {
      "CarOther"
    } else {
      "Unknown"
    })
```

## Data Preparation cd...

```{r , cache=TRUE}
GermanCredit$Purpose <- 
  names(GermanCredit[20:30])[max.col(GermanCredit[20:30])]
```

## Data Preparation cd... {.smaller}

```{r , cache=TRUE}
GermanCredit <- GermanCredit[, c(1:7, 10, 63:65)]
str(GermanCredit, width = 80, strict.width = "wrap")
```

## Data visualization in R

- Using base R functions

- Using the `ggplot2` package $\leftarrow$ our focus today

- Using a variety of other packages like `lattice`, `ggvis`, etc.

## The Grammar of Graphics

- Visualisation concept created by Wilkinson (1999)
    - to define the basic elements of a statistical graphic 

- Adapted for R by Wickham (2009) who created the `ggplot2` package
    - consistent and compact syntax to describe statistical graphics
    - highly modular as it breaks up graphs into semantic components 

- Is not a guide which graph to choose and how to convey information best!

Source: https://rpubs.com/timwinke/ggplot2workshop

## The Grammar of Graphics - Terminology

A statistical graphic is a...

- mapping of **data**
- to **aesthetic attributes** (color, size, xy-position)
- using **geometric objects** (points, lines, bars)
- with data being **statistically transformed** (summarised, log-transformed)
- and mapped onto a specific **facet** and **coordinate system**
    
## Age vs. Amount {.smaller}

```{r echo=FALSE, fig.height=3, fig.width=5, warning=FALSE, cache=T}
ggplot(data = GermanCredit, aes(x = Age, y = Amount))+ geom_point() 
library(magick)
library(grid)
frink <- image_read("https://jeroenooms.github.io/images/frink.png")
homer <- image_read("http://www.logoeps.com/wp-content/uploads/2013/01/bart-simpson-logo-vector1.png")
grid.raster(frink)
```

- Which data is used as an input?
- What geometric objects are chosen for visualization?
- What variables are mapped onto which attributes?
- What type of scales are used to map data to aesthetics?
- Are the variables statistically transformed before plotting?

## Anatomy of ggplots

```
ggplot(data = [dataframe], aes(x = [var_x], y = [var_y], 
       color = [var_for_color], fill = [var_for_fill], 
       shape = [var_for_shape])) +
    geom_[some_geom] +
    ... # other options
```

## Biker age vs. crash hour - code

```{r fig.height=3, fig.width=5}
ggplot(data = GermanCredit, aes(x = Age, y = Amount)) +  geom_point()
```

## Altering features {.smaller}

```{r echo=FALSE, fig.height=3, fig.width=5, warning=FALSE}
ggplot(data = GermanCredit, aes(x = Age, y = Amount)) + geom_point(alpha = 0.5, color = "DarkGreen", 
             shape = 17, size = I(3))
grid.raster(homer)
```

- How did the plot change?
- Are these changes based on data (i.e. can be mapped to variables in the dataset) or
changes in preferences for geometric objects?

## Altering features - code {.smaller}

```{r fig.height=3, fig.width=5}
ggplot(data = GermanCredit, aes(x = Age, y = Amount, size = 2)) +
  geom_point(alpha = 0.5, color = "Sky Blue", shape = 17)
```

## Test yourself 

<div class="question">
How about creating this kind of graph? 
</div>

```{r fig.height=4, fig.width=7, echo = FALSE}
GermanCredit_tmp <- GermanCredit[1:200, ]
ggplot(data = GermanCredit_tmp, aes(x = Amount, y = Duration,
                                color = Sex, size = Duration*Amount)) +
  geom_point(alpha = 0.3)
```

## Test yourself cd..

First step, trim the data.

```{r fig.height=4, fig.width=7}
GermanCredit_tmp <- GermanCredit[1:200, ]
```

## Test yourself cd..

```{r fig.height=4, fig.width=7, echo = FALSE}
GermanCredit_tmp <- GermanCredit[1:200, ]
ggplot(data = GermanCredit_tmp, aes(x = Amount, y = Duration,
                                color = Sex, size = Duration*Amount)) +
  geom_point(alpha = 0.3)
```

## More alterations {.smaller}

```{r echo=FALSE, fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = GermanCredit, aes(x = Age, y = Amount, color = Duration)) +
  geom_point(alpha = 0.45) +
  facet_grid(. ~ Sex)
```

- How did the plot change?
- Are these changes based on data (i.e. can be mapped to variables in the dataset) or
changes in preferences for geometric objects?

## More alterations - code {.smaller}

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = GermanCredit, aes(x = Age, y = Amount, color = Duration)) +
  geom_point(alpha = 0.45) +
  facet_grid(. ~ Sex)
```

## More alterations - code cd... {.smaller}

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = GermanCredit, aes(x = Age, y = Amount, color = Duration,
                                size = factor(NumberExistingCredits))) +
  geom_point(alpha = 0.45) +
  facet_grid(. ~ Sex)
```


## Anatomy of ggplots

```
ggplot(data = [dataframe], aes(x = [var_x], y = [var_y], 
       color = [var_for_color], fill = [var_for_fill], 
       shape = [var_for_shape])) +
    geom_[some_geom] +
    ... # other options
```

## Histograms

```{r fig.height=3, fig.width=7, warning=FALSE}
ggplot(data = GermanCredit, aes(x = Amount)) +
  geom_histogram(binwidth = 500)
```

## Boxplots

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = GermanCredit, aes(y = Age, x = Sex, fill = Sex)) +
  geom_boxplot()
```

## Barplots

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = GermanCredit, aes(x = Property)) +
  geom_bar()
```

## Segmented barplots

```{r fig.height=3, fig.width=9, warning=FALSE}
ggplot(data = GermanCredit, aes(x = Property, fill = Purpose)) +
  geom_bar()
```

## Segmented barplots - proportions

```{r fig.height=3, fig.width=9, warning=FALSE}
ggplot(data = GermanCredit, aes(x = Property, fill = Purpose)) +
  geom_bar(position="fill")
```

## Try yourself 

<div class="question">
How about creating this kind of graph? 
</div>

```{r fig.height=3, fig.width=8, warning=FALSE, echo = FALSE}
k <- ggplot(GermanCredit, aes(Purpose, fill=Class))
k1 <- k + geom_bar(position="dodge") 
k2 <- k1 + scale_fill_brewer() 
k3 <- k2 + coord_flip()
k4 <- k3 + facet_grid(. ~ Sex)
k4
```

## Try yourself cd..

```{r fig.height=3, fig.width=8, warning=FALSE, echo = FALSE}
k1
```

## Try yourself cd..

```{r fig.height=3, fig.width=8, warning=FALSE, echo = FALSE}
k2
```

## Try yourself cd..

```{r fig.height=3, fig.width=8, warning=FALSE, echo = FALSE}
k3
```

## Try yourself DONE

```{r fig.height=3, fig.width=8, warning=FALSE, echo = FALSE}
k4
```

## More `ggplot2` resources

- Visit http://docs.ggplot2.org /current/ for documentation on the current version
of the `ggplot2` package. It's full of examples!

- Refer to the `ggplot2` 
[cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/08/ggplot2-cheatsheet.pdf).

## `plotly` trailer {.smaller}

```{r fig.height=4, fig.width=7, warning=FALSE}
plot_ly(GermanCredit, x = Duration, y = Amount, 
        text = paste("Duration: ", Duration),
        mode = "markers", color = Age, size = Duration)
```

## What else can be done with R 

```{r fig.height=5, fig.width=7, warning=FALSE, echo = FALSE}
# Load package
library(networkD3)

# Load data
data(MisLinks)
data(MisNodes)

# Plot
forceNetwork(Links = MisLinks, Nodes = MisNodes,
            Source = "source", Target = "target",
            Value = "value", NodeID = "name",
            Group = "group", opacity = 0.8,
            zoom = TRUE)
```

## What else can be done with R {.smaller}

```{r fig.height=5, fig.width=7, warning=FALSE, echo = FALSE}
suppressMessages(library(leaflet))

m <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=44.7693005, lat=41.6881125, popup="Techpark Tbilisi")
m
```

# Data wrangling

## Data wrangling in R

- Using base R functions

- Using the `dplyr` package $\leftarrow$ our focus today

- Using a variety of other packages like `plyr`, `tidyr`, `lubridate`, etc.

## Data wrangling with `dplyr`

The `dplyr` package is based on the concepts of functions as verbs that manipulate data frames:

- `filter()`: pick rows matching criteria
- `select()`: pick columns by name 
- `arrange()`: reorder rows 
- `mutate()`: add new variables
- `summarise()`: reduce variables to values

## `dplyr` rules

- First argument is a data frame
- Subsequent arguments say what to do with data frame
- Always return a data frame 
- Avoid modify in place

## Filter rows with `filter()`

- Select a subset of rows in a data frame.
- Easily filter for many conditions at once.

## `filter()` {.smaller}

for credits for new car

```{r}
GermanCredit %>%
  filter(Purpose == "Purpose.Retraining") %>% 
  select(Age, Amount, Sex, Class, Property)
```

## `filter()` {.smaller}

Credits with purpose Retraining just for males
 
```{r}
GermanCredit %>%
  filter(Purpose == "Purpose.Retraining", Sex == "Male") %>% 
  select(Age, Amount, Sex, Class, Property)
```

## Commonly used logical operators in R {.smaller}

operator    | definition
------------|--------------------------
`<`         | less than
`<=`        |	less than or equal to
`>`         | greater than
`>=`        |	greater than or equal to
`==`        |	exactly equal to
`!=`        |	not equal to
`x | y`     | `x` OR `y`
`x & y`     | `x` AND `y`

## Commonly used logical operators in R {.smaller}

operator     | definition
-------------|--------------------------
`is.na(x)`   | test if `x` is `NA`
`!is.na(x)`  | test if `x` is not `NA`
`x %in% y`   | test if `x` is in `y`
`!(x %in% y)`| test if `x` is not in `y`
`!x`         | not `x`

## Aside: real data is messy! {.smaller}

<div class="question">
This format of variable is not to nice, what can we do about it?
</div>

```{r}
GermanCredit %>%
  group_by(Purpose) %>%
  summarise(Purpose_count = n())
```

## Cleaning up the data!

- We're going to need to do some text parsing to clean up
these data
  
- New R package: `stringr`

## Install and load: `stringr`

- Install:
```{r eval=FALSE}
install.packages("stringr")
```

- Load:
```{r}
library(stringr)
```

- Package reference: Most R packages come with a vignette that describe
in detail what each function does and how to use them, they're incredibly
useful resources (in addition to other worked out examples on the web)
https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html

## Replace with `str_replace()` and add new variables with `mutate()` {.smaller}

- Remember we want to do the following in the `Purpose` variable: Delete Purpose prefix for all the values
 
```{r}
GermanCredit <- GermanCredit %>%
  mutate(Purpose = str_replace(Purpose, "Purpose.", "")) 
```

- Note that we're overwriting existing data and columns, so be careful!
    + But remember, it's easy to revert if you make a mistake since we didn't
    touch the raw data, we can always reload it and start over

## Check before you move on {.smaller}

Always check your changes and confirm code did what you wanted it to do

```{r}
GermanCredit %>%
  group_by(Purpose) %>%
  summarise(Purpose_count = n()) 
```

## `slice()` for certain row numbers {.smaller}

First five
```{r}
GermanCredit %>%
  slice(1:5)
```

## `slice()` for certain row numbers {.smaller}

Last five
```{r}
last_row <- nrow(GermanCredit)
GermanCredit %>%
  slice((last_row-4):last_row)
```

## `select()` to keep only the variables you mention {.smaller}

```{r}
GermanCredit %>%
  select(Class, Sex) %>%
  table()
```

## or `select()`to exclude variables {.smaller}

```{r}
GermanCredit %>%
  select(-Property) %>%
  slice(1:5)
```

## `rename()` specific columns {.smaller}

Correct typos and rename to make variable names shorter and/or more informative

- Original names:
```{r}
names(GermanCredit)
```

- Rename `Sex` to `Gender`:
```{r}
GermanCredit <- GermanCredit %>%
  rename(Gender = Sex)
```

## Check before you move on {.smaller}

Always check your changes and confirm code did what you wanted it to do

```{r}
names(GermanCredit)
```

## `summarise()` in a new data frame {.smaller}

```{r}
GermanCredit %>%
  group_by(Purpose) %>%
  summarise(count = n()) %>%
  arrange(count)
```

## and `arrange()` to order rows {.smaller}

```{r}
GermanCredit %>%
  group_by(Purpose) %>%
  summarise(count = n()) %>%
  arrange(desc(count))
```

## More `dplyr` resources

- Visit https://cran.r-project.org/web/packages/dplyr/vignettes/introduction.html for the
package vignette.

- Refer to the `dplyr` 
[cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf).

# Basic R syntax

## Few important R syntax notes

For when not working with `dplyr` or `ggplot2`

- Refer to a variable in a dataset as `GermanCredit$Class`

- Access any element in a dataframe using square brackets

```{r}
GermanCredit <- as.data.frame(GermanCredit)
GermanCredit[1,5] # row 1, column 5
```
    
    - For all observations in row 1: `GermanCredit[1, ]`
    - For all observations in column 5: `GermanCredit[, 5]`

# What's next?

## Want more R? {.smaller}

- Local install
    - R: https://cran.r-project.org/
    - RStudio: https://www.rstudio.com/products/RStudio/#Desktop

- Resources for learning R:
    - [Coursera](https://www.coursera.org/)
    - [DataCamp](https://www.datacamp.com/)
    - Many many online demos, resources, examples, as well as books 

- Debugging R errors:
    - Read the error!
    - [StackOverflow](http://stackoverflow.com/)

- Keeping up with what's new in R land:
    - [R-bloggers](http://www.r-bloggers.com/)
    - Twitter: #rstats

## More materials 

- [List of useful packages](https://support.rstudio.com/hc/en-us/articles/201057987-Quick-list-of-useful-R-packages)

- [Shiny tutorial](http://shiny.rstudio.com/tutorial/lesson1/)

- [Cheatsheets](https://www.rstudio.com/resources/cheatsheets/)

- [ggplot2 compendium](http://docs.ggplot2.org/current/) 

- [dplyr tutorial](https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a)

- [ggplot2 tutorial](http://www.education.umd.edu/EDMS/fac/Harring/769L-Spring-2016/Resources/R-Resources/ggplot2-tutorial.pdf)

- [visualizations htmlwidgets](http://gallery.htmlwidgets.org/)

- [visualizations metricgraphics](http://hrbrmstr.github.io/metricsgraphics/)

- [visualizations plotly](https://plot.ly/r/)

## Useful Georgian read.csv

- [Github page Jean-Sebastien Gourdet](https://github.com/geojsg/GeorgianUnicode)

# Questions 

# Exercise

## Your turn
Create a new dataframe that doesn't include observations where `Purpose = Other` since 
it's not clear what this means.

This new dataframe also should include observations for Man only, and for people older than 40 and younger than 60 

Summarize the number of Good and Bad credits for each grup of `Property` arrange the table descending order.

Create a visualization that will help answer which Married status is the most probable to have bad credits for whole data set. 

## Contact 

- Janek Idziak Jan.Idziak@gmail.com 

- Michal Krason michal.krason1@gmail.com 