---
title: "Using R"
author: "Michal Krason"
date: "September 10, 2016"
output: ioslides_presentation
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Data structures
## Basic data types
Four most used atomic (lowest level) data types are:

* Logical: TRUE/FALSE
```{r}
is.logical(FALSE)
```
* Integer: 1, 2, -3 etc.
```{r}
is.integer(1L)
```
https://github.com/jandziak/Tbilisi_R_intro

## Basic data types

* Double: 1.3, 4.2, -3 etc.
```{r}
is.double(1.7)
```
* Character: abcm, t, random
```{r}
is.character("random")
```
## Basic data types

We can also check if our data is numeric, i.e. double or integer
```{r}
is.numeric(1L)
is.numeric(3.5)
```
Converting the values between types is possible in the order they were presented: logical -> integer -> double -> character.

This will be important e.g. when combining vectors (coercion).

## Assigning variables

Variables are assigned using '<-' operator.
```{r}
a <- "Just for show"
a
b <- c(7,4,5454)
b[3]
```

## Vectors

* Contain only atomic type elements

* All elements are of the same type, different are coerced according to the priority presented before

```{r}
c(1L, 2.3)
c(TRUE, "Take that!")
```

## Exercise

* Predict the coercion of the following vectors
```{r, eval=FALSE}
c(1, FALSE)
c("a", 1)
c(TRUE, 1L)
c(FALSE, "abc")
```

## Creating vectors

* Shortcut for defining sequence
```{r}
2:6
```
* More complex sequence
```{r}
seq(2, 3, by=0.5)
```

## Basic functions

* Repeating elements
```{r}
rep(2:4, times=3)
x<-rep(c(7,3,6), times=c(1,4,3)); x
```
* Sorting
```{r}
sort(x)
```

## Basic functions

* Reversing vector order
```{r}
rev(x)
```
* Summarizing amounts of elements
```{r}
table(x)
```
## Basic functions
* Checking the class of the element
```{r}
class(x)
```
* Checking the structure of the element
```{r}
str(x)
```

## Selecting elements: by position
* The fourth element
```{r}
x[4]
```
* All but the fourth.
```{r}
x[-4] 
```
## Selecting elements: by position
* Elements two to four.
```{r}
x[2:4] 
```
* Elements one and five.
```{r}
x[c(1, 5)]
```
## Selecting elements: by value
* Elements which are equal to 7.
```{r}
x[x == 7]
```
* All elements bigger than three.
```{r}
x[x > 3]
```
## Selecting elements: by value
* Elements in the set 3, 6.
```{r}
x[x %in% c(3, 6)]
```
## Matrixes
Let's create a vector of observations drawn from normal distribution and turn it into a matrix!
```{r}
x <- rnorm(12)
m <- matrix(x, nrow = 3, ncol = 4)
m
```
## Matrixes: selecting elements
We can choose a single row
```{r}
m[1, ]
```
or a single column.
```{r}
m[, 3]
```
They are automatically turned into a vector. Transposing matrices is done using t().

## Lists
* Can contain different atomic types
* Can be nested - a list may consist of several lists
```{r}
l <- list("a" = c(1,2,3), "bla", list(c(TRUE,FALSE), c(1L, 3L))); l
```
## Lists: selecting elements
* Selecting an element of a list
```{r}
l[[1]]
```
* Selecting a list containing an element - preserve structure
```{r}
l[2]
```
## Lists: selecting elements
* Selecting a named element
```{r}
l$a
```
* Addressing nested elements
```{r}
l[[3]][[2]]
```
## Data frames
* A special case of a list where all elements are the same length.
```{r}
df <- data.frame(x = 1:3, y = c('a','b','c')); df
```


## Data frames: selecting elements
* Can be accessed as a matrix
```{r}
df[, 1]
```
* or as a list.
```{r}
df$y
```
Something weird happened to 'y'. This is important, to avoid it we should set stringsAsFactors = FALSE.

# Elements of syntax
## Commonly used logical operators in R {.smaller}

operator    | definition
------------|--------------------------
`<`         | less than
`<=`        |	less than or equal to
`>`         | greater than
`>=`        |	greater than or equal to
`==`        |	exactly equal to
`!=`        |	not equal to
`x | y`     | `x` OR `y`
`x & y`     | `x` AND `y`

## Commonly used logical operators in R {.smaller}

operator     | definition
-------------|--------------------------
`is.na(x)`   | test if `x` is `NA`
`!is.na(x)`  | test if `x` is not `NA`
`x %in% y`   | test if `x` is in `y`
`!(x %in% y)`| test if `x` is not in `y`
`!x`         | not `x`

## Exercise
* Load GermanCredit data using data()
* Select every second row
* Remove the rows that have maximum Duration variable
* Select only Duration, Amount, Age and Telephone variables
* Remove the fourth variable

## Exercise
```{r, eval=FALSE}
data("GermanCredit")
exercise1 <- GermanCredit[seq(1,length(GermanCredit[,1]),2),]
exercise1 <- exercise1[exercise1$Duration != max(exercise1$Duration),]
exercise1 <- exercise1[c("Duration", "Amount", "Age", "Telephone")]
exercise1 <- exercise1[, -4]
```

## If statements

* Below an example of an if statement
```{r}
i <- 3
if (i > 3){
  print("Yes")
} else if (i == 3){
  print("No")
} else {
  print("I don't know")
}
```
## For loop

* Below an example of a for loop
```{r}
for (i in 1:4){
  j <- i + 10
  print(j)
}
```
## While loop
* Below an example of a while loop
```{r}
i <- 1
while (i < 5){
  print(i)
  i <- i + 1
}
```
## Functions
* Below an example of a function 
```{r}
square <- function(x){
  squared <- x*x
  return(squared)
}
```
## Exercise

* Write a function that for given size produces a multiplication table of that size
* For loop is probably the easiest way to do this

## Exercise
```{r}
mult_table <- function(size) {
  m <- matrix (0, ncol = size, nrow = size)
  for (i in 1:size) {
    for (j in 1:size) {
      m[i,j] <- i*j
    }
  }
  return(m)
}
mult_table(6)
```

# Useful functions
## Useful functions: math

* abs(x)
* round(x, n)
* log(x) 
* sum(x)
* exp(x)
* mean(x)
* sd(x)
* quantile(x, q - optional)
* cor(x, y)

## Useful functions: reading and writing data
```{r echo = F, warning=F, message=F}
library(caret)
setwd("C:\\Repository\\R_workshop\\Tbilisi_R_intro\\Michal")
```

```{r}
GermanCredit <- read.table("GermanCredit.txt") 
GermanCredit2 <- read.csv("GermanCredit2.csv") 
str(GermanCredit)
GermanJoint <- cbind(GermanCredit, GermanCredit2)
write.csv(GermanJoint, "GermanJoint.csv")
```
## Useful functions: strings
* Join multiple vectors together.
```{r}
x <- 1:5
y <- rev(x)
paste(x, y, sep = '')
```
* Join elements of a vector together.
```{r}
paste(x, collapse = ' ') 
```
## Useful functions: strings
* Find regular expression matches in x.
```{r}
x <- c("random.mail@gmail.com", "bla", "abc@d", "ef@gh.i")
pattern <- "@([a-z]+)\\."
grep(pattern, x) 
```
* Replace matches in x with a string.
```{r}
gsub(pattern, "@spam.", x)
```
## Useful functions: strings
* Convert to uppercase / lowercase.
```{r}
x <- "aBvdaET"
toupper(x)
tolower(x)
```
## Useful functions: linear model
* Remember the german data? Let's look for linear dependencies.
```{r}
lm(Amount ~ Duration, data = GermanJoint)
```
## Useful functions: basic plot
Now let's see it on a plot.
```{r}
plot(GermanJoint$Duration, GermanJoint$Amount)
```

## Useful functions: sapply

Instead of using loops, in many cases it is faster to work on vectors using 'apply' family.
```{r}
m<-matrix(rnorm(1000),ncol=5,nrow=200)
sapply(c(1:ncol(m)),function(i) {mean(m[ ,i])})
```
## Exercise
* Read in German data part 3
* Join into one data frame
* Extract the names containing "credit" (regardless of capitalization)
* Use apply to calculate the standard deviation of each of those columns
* Save the result in a csv file

## Exercise
```{r}
GermanCredit3 <- read.csv("GermanCredit3.csv")
GermanFinal <- cbind(GermanJoint, GermanCredit3)
GermanSel <- GermanFinal[, grep("credit",tolower(names(GermanFinal)))]
GermanSD <- sapply(c(1:length(GermanSel[1,])),
                   function(i) {sd(GermanSel[, i])})
GermanSD
write.csv(GermanSD, "Result.csv")
```

## Libraries in R

* Download and install a package from CRAN.
```{r, eval=FALSE}
install.packages(‘dplyr’)
```
* Load the package into the session, making all
its functions available to use.
```{r, eval=FALSE}
library(‘dplyr’)
```
* Use a particular function called explicitly from a package.
```{r, eval=FALSE}
dplyr::select
```
